Engineering infinite-range SU($n$) interactions with spin-orbit-coupled fermions in an optical lattice

### 2022-10-04 sun_phases/spin_model_paper/main.tex
updating spin model paper materials

%%% symbols, notations, etc.
\usepackage{physics,braket,bm,amssymb} % physics and math
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}[2]{\dfrac{#1}{#2}} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\braket} % shorthand for braket notation
\newcommand{\Bk}{\Braket} % shorthand for braket notation
\renewcommand{\v}{\bm} % bold vectors
\renewcommand{\d}{\partial} % partial derivative
\renewcommand{\dd}{\text{d}} % for infinitesimals
\renewcommand{\i}{\mathrm{i}\mkern1mu} % imaginary unit

% double angle brackets
\newcommand{\bbk}[1]{\langle\!\langle #1 \rangle\!\rangle}
\newcommand{\Bbk}[1]
{\left\langle\!\!\left\langle #1 \right\rangle\!\!\right\rangle}

@@

% symbols with hats
\renewcommand{\b}{\hat b}
\renewcommand{\c}{\hat c}
\newcommand{\s}{\hat s}
\renewcommand{\H}{\hat H}
\renewcommand{\S}{\hat S}
\newcommand{\R}{\hat R}
\newcommand{\T}{\hat T}
\renewcommand{\O}{\hat{\mathcal{O}}}
\renewcommand{\P}{\hat{\mathcal{P}}}

@@

\newcommand{\xx}{\x\x}
\newcommand{\xxi}{\x\x_\i}
\newcommand{\X}{\text{X}}
\newcommand{\XX}{\X\X}
\newcommand{\XXI}{\X\X_\i}
\newcommand{\spin}{\text{spin}}
\newcommand{\eff}{\text{eff}}
\newcommand{\crit}{\text{crit}}
\newcommand{\MF}{\text{MF}}
\newcommand{\su}{\mathfrak{su}}
\newcommand{\sds}{\bar{\v s}\cdot\bar{\v s}}
\DeclareMathOperator{\diag}{diag}

@@

%%% figures
\usepackage{graphicx} % for figures
\graphicspath{{./figures/}} % set path for all figures

@@

\date{22 September 2021}

@@

SU($n$) symmetries play an important role in physics.
Underpinning much of high energy physics, the SU($n$) gauge theory known as Yang-Mills theory is central to our understanding of the electroweak and strong forces.
Extensions of Yang-Mills and SU($n$) symmetry feature in the most well-studied examples of holographic duality \cite{maldacena1999largen} and the connection between entanglement and gravity \cite{ryu2006holographic} through the anti-de Sitter/conformal field theory (AdS/CFT) correspondence.
In a condensed matter setting, SU(2) appears ubiquitously as a symmetry of the Hubbard model, with important consequences for the study of quantum magnetism and high temperature superconductivity \cite{lee2006doping}.
The extension of SU(2) Hubbard and spin models to SU($n$) has led to predictions of exotic phases of matter such as valence bond solids \cite{read1989valencebond, rokhsar1990quadratic, kaul2012lattice, hermele2011topological} and chiral spin liquids \cite{hermele2009mott, hermele2011topological, chen2016syntheticgaugefield, nataf2016chiral}, as well as the potential to perform universal topological quantum computation \cite{freedman2004class, nayak2008nonabelian} and other phenomena \cite{nataf2014exact, nataf2016exact}.
Furthermore, disordered SU($n$) spin models have opened analytically tractable avenues for studying quantum chaos and information scrambling \cite{sachdev1993gapless}.

The tremendous theoretical significance of SU($n$) symmetries makes it all the more exciting that they appear naturally in experimental atomic, molecular, and optical (AMO) platforms with exquisite degrees of microscopic control.
This symmetry arises through the independence of atomic orbital and interaction parameters on the $n$ nuclear spin states of alkaline-earth(-like) atoms, with e.g.~$n=10$ for ${}^{87}$Sr \cite{wu2003exact, cazalilla2009ultracold, gorshkov2010twoorbital, cazalilla2014ultracold}.
As a result, AMO experiments can directly probe the role of SU($n$) interactions in controllable settings.
Recent progress includes studies of the thermodynamic properties of SU($n$) fermionic gases \cite{hazzard2012hightemperature, bonnes2012adiabatic, stellmer2013degenerate, yip2014theory, pagano2014onedimensional, choudhury2020collective, song2020evidence, sonderhouse2020thermodynamics}, SU($n$) Hubbard phases and phase transitions \cite{taie2012su, hofrichter2016direct, taie2020observation}, single- \cite{messio2012entropy} and two-orbital \cite{cappellini2014direct, scazza2014observation, zhang2014spectroscopic, beverland2016realizing} SU($n$) magnetism, and multi-body SU($n$)-symmetric interactions \cite{goban2018emergence, perlin2019effective}.

@@

Here we derive a collective SU($n$) spin model for a system of ultracold alkaline-earth(-like) atoms trapped in an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by the single-body kinetic and two-body interaction Hamiltonians
\begin{align}
  \H_{\t{kin}}
  &= -J \sum_{\bk{j,j'},\mu} \c_{j\mu}^\dag \c_{j'\mu} + \t{h.c.},
  \label{eq:H_kin_FH} \\
  \H_{\t{int}}
  &= \f{U}{2} \sum_{j,\mu,\nu}
  \c_{j\mu}^\dag \c_{j\mu} \c_{j\nu}^\dag \c_{j\nu},
  \label{eq:H_int_FH}
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $\c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude (for simplicity assumed to be the same in all directions); and $U$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which is a good approximation for a sufficiently deep lattice, namely when $J\lesssim E_{\t{R}}$, where $E_{\t{R}}$ is the atom recoil energy.
For simplicity, we now assume a one-dimensional periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$ (in units with lattice spacing $a=1$), $\c_{j\mu} = \frac1{\sqrt{L}} \sum_q e^{-\i q\cdot j} \c_{q\mu}$, finding that
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \c_{q\mu}^\dag \c_{q\mu}, \\
  \H_{\t{int}}
  &= \f{u}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  \c_{k\mu}^\dag \c_{\ell\mu} \c_{p\nu}^\dag \c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $u\equiv U\times N/L$ for convenience, $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and zero otherwise (enforcing conservation of momentum).

If the interaction energy $U$ is small compared to the single-particle bandwidth $4J$, then the mode-changing collisions in $\H_{\t{int}}$ become off-resonant, motivating the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (i.e.~either $k=\ell$ and $p=q$, or $k=q$ and $p=\ell$)\footnote{Note that the frozen-mode approximation neglects correlated momentum-hopping terms of the form $\c_{\pi-p,\mu}^\dag \c_{\pi-q,\mu} \c_{p\nu}^\dag \c_{q\nu}$, which conserve both momentum and energy.
We defer a careful treatment of these terms to future work, noting only that they vanish on the manifold of permutationally symmetric spin states with one atom per lattice site, and that the frozen-mode approximation is benchmarked in Refs.~\cite{he2019engineering, smale2019observation} and Appendix \ref{sec:benchmarking}.}.
The terms with $k=\ell$ and $p=q$ are $\frac{u}{2N} \sum \c_{k\mu}^\dag \c_{k\mu} \c_{p\nu}^\dag \c_{p\nu} = \frac12 N u$, which is a constant  energy shift that we can freely neglect.
Defining the spin operators $\s_{\mu\nu q}\equiv \c_{q\mu}^\dag \c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians are
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \s_{\mu\mu q},
  \label{H_kin_start} \\
  \H_{\t{int}}
  &= -\f{u}{2N} \sum_{p,q,\mu,\nu} \s_{\mu\nu p} \s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by fermionic statistics (Pauli exclusion).
In this case we can simply treat our system as $N$ distinguishable $n$-level quantum spins at ``lattice sites'' $p,q$.
Note that the ``kinetic'' terms of this spin model ($\H_{\t{kin}}$) are proportional to the identity operator, contributing an overall shift in energy that we can neglect at this point.
Nevertheless, these kinetic terms will become important in the presence of an external drive, which we discuss in Section \ref{sec:SOC}.
The validity of approximating the Hubbard model in Eqs.~\eqref{eq:H_kin_FH}--\eqref{eq:H_int_FH} by the spin model in Eqs.~\eqref{H_kin_start}--\eqref{eq:H_int_start} has been previously benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}, and we provide additional benchmarking for SU(4) and SU(6) in Appendix \ref{sec:benchmarking}.

To further simplify the interaction Hamiltonian $\H_{\t{int}}$ and write it in a form reminiscent of more familiar SU(2) spin models, we now construct the operator-valued spin matrix
\begin{align}
  \v\s_q \equiv \sum_{\mu,\nu} \s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\v{\hat{A}},\v{\hat{B}}$, we define the inner product
\begin{align}
  \v{\hat{A}} \cdot \v{\hat{B}}
  \equiv \sum_{\mu,\nu} \hat A_{\mu\nu}^\dag \hat B_{\mu\nu}.
  \label{eq:dot_product}
\end{align}
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  \H_{\t{int}} = -\f{u}{2N} \sum_{p,q} \v\s_p\cdot\v\s_q
  = -\f{u}{2N}\v\S\cdot\v\S,
  \label{eq:H_int}
\end{align}
where $\v\S\equiv\sum_q\v\s_q$ is a collective spin matrix, analogous to the collective spin vector $\vec S=(\S_\x,\S_\y,\S_\z)$ in the case of SU(2) \cite{he2019engineering}, with $\frac12\v\S\cdot\v\S \simeq \vec S\cdot\vec S = \S_\x^2 + \S_\y^2 + \S_\z^2$ when $n=2$ (here $\simeq$ denotes equality up to identity terms).

\begin{figure}
\centering
\includegraphics[width=\linewidth]{spin_model.pdf}
\caption{
(a) Ultracold atoms on a lattice of $L$ sites tunnel between neighboring lattice sites at a rate $J$, and locally repel each other with interaction energy $U$.
(b,c) When the interaction energy $U$ is small compared to the single-particle bandwidth $4J$, the frozen-mode approximation enables the interaction Hamiltonian to be written as a spin model consisting of exchange terms $\v\s_p\cdot\v\s_q$, which swap the states of two spins pinned to modes $p,q$.
(d) Interactions open an energy gap $u=U\times N/L$ between the manifold of permutationally symmetric states of $N$ spins, and the orthogonal complement of states that break spin-permutation symmetry.
}
\label{fig:spin_model}
\end{figure}

We now discuss the spin Hamiltonian $\H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v\s_p\cdot\v\s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v\s_p\cdot\v\s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v\s_p\cdot\v\s_q$ is analogous to the enforcement of SU(2) spin alignment by ferromagnetic interactions, which similarly assigns different energies to the anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.
By summing over all pair-wise exchange terms $\v\s_p\cdot\v\s_q$, the interaction Hamiltonian $\H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $u$ between the manifold of all permutationally symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
See Figure \ref{fig:spin_model} for a summary of this section thus far.

@@

Finally, we take a moment to discuss individual $n$-level spins.
The state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of an $n$-level spin can be represented by a quasi-probability distribution $Q_\psi$ on the Bloch sphere (commonly known as the Husimi-$Q$ function, e.g.~in the spin-squeezing community \cite{ma2011quantum}).
The value $Q_\psi\p{\v v}$ at a point $\v v$ on the sphere is equal to the overlap of $\ket\psi$ with a pure state $\ket{\v v}$ that is maximally polarized in the direction of $\v v$: $Q_\psi\p{\v v}\equiv\abs{\bk{\v v|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\hat\rho$, this distribution is defined by $Q_{\hat\rho}\p{\v v}\equiv\bk{\v v|\hat\rho|\v v}$.
Closely related spherical representations of multilevel spin states and operators are discussed in Refs.~\cite{dowling1994wigner, li2013weylwignermoyal}.
In practice, it is conceptually useful to identify the Hilbert space of a single $n$-level spin with the Dicke manifold of $n-1$ spin-$\frac12$ particles.

@@

We consider a specific three-laser driving scheme with a geometry sketched in Figure \ref{fig:3LD}.
Here the lattice lies in the $y$-$z$ plane at an angle $\theta$ to the $z$ axis, oriented along $\v\ell=(0,\sin\theta,\cos\theta)$.
We set the spin quantization axis along $z$.
The laser setup consists of
\begin{enumerate*}
\item two counter-propagating right-circularly polarized lasers with drive amplitudes $\Omega_\pm$ and wavevectors $\kappa\v v_\pm$, propagating in opposite directions along the $z$ axis, $\v v_\pm=(0,0,\pm1)$, and
\item a third laser linearly polarized along $z$, with drive amplitude $\Omega_0$ and wavevector $\kappa \v v_0$, propagating along the $x$ axis $\v v_0=(1,0,0)$.
\end{enumerate*}
All driving lasers are detuned by $\Delta$ below an electronic transition.
The full Hamiltonian for this three-laser drive can be written as
\begin{align}
  \H_{\t{3LD}}^{\t{full}}
  = \sum_{j,m} \Omega_m
  \p{e^{-\i m\phi j} \s_{mj} \otimes\op{\e}{\g}_j + \t{h.c.}}
  + \Delta \hat N_\e,
\end{align}
where $m\in\set{+1,0,-1}$ indexes the laser pointing along $\v v_m$; the SOC angle $\phi\equiv\kappa\v v_+\cdot\v\ell=\kappa \cos\theta$ (in units with lattice spacing $a=1$); $\s_{\z,j},\s_{+,j}\s_{-,j}$ are standard axial, spin-raising, and spin-lowering operators for the spin at lattice site $j$; $\s_{0,j}\equiv\s_{\z,j}$ for shorthand; $\ket\g_j$ and $\ket\e_j$ respectively denote the ground and excited electronic states of atom $j$; and $\hat N_\e=\1\otimes\sum_j\op\e_j$ counts the number of excited atoms (with $\1$ the identity operator on all spin degrees of freedom).

@@

We thank Victor Gurarie, Emil Yuzbashyan, Asier P.~Orioli, and Jeremy T.~Young for helpful discussions on this work.
This work was supported by the AFOSR grants FA9550-18-1-0319, FA9550-19-1-027, by the DARPA and ARO grant W911NF-16-1-0576, ARO W911NF-19-1-0210, DOE (QSA), NSF PHY1820885, NSF JILA-PFC PHY-1734006, NSF QLCI-2016244, and by NIST.

>>>>

%%% symbols, notations, etc.
\usepackage{physics,braket,bm,amssymb} % physics and math
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}[2]{\dfrac{#1}{#2}} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\braket} % shorthand for braket notation
\newcommand{\Bk}{\Braket} % shorthand for braket notation
\renewcommand{\d}{\partial} % partial derivative
\renewcommand{\dd}{\text{d}} % for infinitesimals
\renewcommand{\i}{\mathrm{i}\mkern1mu} % imaginary unit

% double angle brackets
\newcommand{\bbk}[1]{\langle\!\langle #1 \rangle\!\rangle}
\newcommand{\Bbk}[1]{\left\langle\!\!\left\langle #1 \right\rangle\!\!\right\rangle}

@@

% symbols with hats
\renewcommand{\b}{\hat{b}}
\renewcommand{\c}{\hat{c}}
\newcommand{\s}{\hat{s}}
\renewcommand{\H}{\hat{H}}
\renewcommand{\S}{\hat{S}}
\newcommand{\R}{\hat{R}}
\newcommand{\T}{\hat{T}}
\renewcommand{\O}{\hat{\mathcal{O}}}
\renewcommand{\P}{\hat{\mathcal{P}}}

@@

\newcommand{\xx}{{\x\x}}
\newcommand{\xxi}{{\x\x_\i}}
\newcommand{\X}{\text{X}}
\newcommand{\XX}{{\X\X}}
\newcommand{\XXI}{{\X\X_\i}}
\newcommand{\spin}{\text{spin}}
\newcommand{\eff}{\text{eff}}
\newcommand{\crit}{\text{crit}}
\newcommand{\MF}{\text{MF}}
\newcommand{\su}{\mathfrak{su}}
\newcommand{\sds}{\bar{\bm{s}}\cdot\bar{\bm{s}}}
\DeclareMathOperator{\diag}{diag}

@@

%%% figures
\usepackage{graphicx} % for figures
\graphicspath{{./figures/}} % set path for all figures
\usepackage[caption=false]{subfig} % subfigures (via \subfloat[]{})
\newcommand{\sref}[1]{\protect\subref{#1}} % for referencing subfigures

@@

\date{15 December 2021}

@@

SU($n$) symmetries play an important role in physics.
Underpinning much of high energy physics, the SU($n$) gauge theory known as Yang-Mills theory is central to our understanding of the electroweak and strong forces.
Extensions of Yang-Mills and SU($n$) symmetry feature in the most well-studied examples of holographic duality \cite{maldacena1999largen} and the connection between entanglement and gravity \cite{ryu2006holographic} through the anti-de Sitter / conformal field theory (AdS/CFT) correspondence.
In a condensed matter setting, SU(2) appears ubiquitously as a symmetry of the Hubbard model, with important consequences for the study of quantum magnetism and high temperature superconductivity \cite{lee2006doping}.
The extension of SU(2) Hubbard and spin models to SU($n$) has led to predictions of exotic phases of matter such as valence bond solids \cite{read1989valencebond, rokhsar1990quadratic, kaul2012lattice, hermele2011topological} and chiral spin liquids \cite{hermele2009mott, hermele2011topological, chen2016syntheticgaugefield, nataf2016chiral}, novel magnetic behaviors \cite{nataf2014exact, nataf2016exact}, and flavor selective phase transitions \cite{xu2018interaction, delre2018selective, tusi2021flavourselective, ibarra-garcia-padilla2021universal} as well as the potential to perform universal topological quantum computation \cite{freedman2004class, nayak2008nonabelian}.
Furthermore, disordered SU($n$) spin models have opened analytically tractable avenues for studying quantum chaos and information scrambling \cite{sachdev1993gapless}.

The tremendous theoretical significance of SU($n$) symmetries makes it all the more exciting that they appear naturally in experimental atomic, molecular, and optical (AMO) platforms with exquisite degrees of microscopic control.
This symmetry arises through the independence of atomic orbital and interaction parameters on the $n$ nuclear spin states of alkaline-earth(-like) atoms, with e.g.~$n=10$ for ${}^{87}$Sr \cite{wu2003exact, cazalilla2009ultracold, gorshkov2010twoorbital, cazalilla2014ultracold}.
As a result, AMO experiments can directly probe the role of SU($n$) interactions in controllable settings.
Recent progress includes studies of the thermodynamic properties of SU($n$) fermionic gases \cite{hazzard2012hightemperature, bonnes2012adiabatic, stellmer2013degenerate, yip2014theory, ibarra-garcia-padilla2021universal, pagano2014onedimensional, choudhury2020collective, song2020evidence, sonderhouse2020thermodynamics}, SU($n$) Hubbard phases and phase transitions \cite{taie2012su, hofrichter2016direct, taie2020observation}, single- \cite{messio2012entropy} and two-orbital SU($n$) magnetism \cite{cappellini2014direct, scazza2014observation, zhang2014spectroscopic, beverland2016realizing}, and multi-body SU($n$)-symmetric interactions \cite{goban2018emergence, perlin2019effective}.

@@

Here we derive a collective SU($n$) spin model for a system of ultracold alkaline-earth(-like) atoms trapped in an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by the single-body kinetic and two-body interaction Hamiltonians
\begin{align}
  \H_{\t{kin}}
  &= -J \sum_{\bk{j,j'},\mu} \c_{j\mu}^\dag \c_{j'\mu} + \t{h.c.},
  \label{eq:H_kin_FH} \\
  \H_{\t{int}}
  &= \f{U}{2} \sum_{j,\mu,\nu}
  \c_{j\mu}^\dag \c_{j\mu} \c_{j\nu}^\dag \c_{j\nu},
  \label{eq:H_int_FH}
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $\c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude (for simplicity assumed to be the same in all directions); and $U$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which is a good approximation for a sufficiently deep lattice, namely when $J\lesssim E_{\t{R}}$, where $E_{\t{R}}$ is the atom recoil energy.
For simplicity, we now assume a one-dimensional periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$ (in units with lattice spacing $a=1$), $\c_{j\mu} = \frac1{\sqrt{L}} \sum_q e^{-\i q\cdot j} \c_{q\mu}$, finding that
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \c_{q\mu}^\dag \c_{q\mu}, \\
  \H_{\t{int}}
  &= \f{u}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  \c_{k\mu}^\dag \c_{\ell\mu} \c_{p\nu}^\dag \c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $u\equiv U\times N/L$ for convenience, and the Kronecker delta $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and zero otherwise (enforcing conservation of momentum).

Momentum-space modes are delocalized across the lattice, which is why interactions between these modes are $O(U/L)=O(u/N)$.
Mode-changing collisions thereby generally become off-resonant when these interactions are weak compared to the single-particle mode spacing, i.e.~$U/L\ll J/L$ (equivalently $U\ll J$)\footnote{
A reasonable objection to this argument is that each momentum-space atom interacts with $O(N)$ other atoms, which suggests that its total interaction energy is $O(U)\gg J/L$.
However, the \textit{typical} energy difference between two states that are coupled by interactions is $O(J)$, so if $U\ll J$ then again interactions become off-resonant in the ``typical'' case, leaving only the ``atypical'' terms that approximately conserve energy, which again leads to the frozen-mode approximation.
}.
In this regime, we can make the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (i.e.~either $k=\ell$ and $p=q$, or $k=q$ and $p=\ell$)\footnote{
Note that the frozen-mode approximation neglects correlated momentum-hopping terms of the form $\c_{\pi-p,\mu}^\dag \c_{\pi-q,\mu} \c_{p\nu}^\dag \c_{q\nu}$, which also conserve energy.
We defer a careful treatment of these terms to future work, noting only (i) that they vanish on the manifold of permutationally symmetric spin states with one atom per lattice site, and (ii) that the frozen-mode approximation is benchmarked in Refs.~\cite{he2019engineering, smale2019observation} and Appendix \ref{sec:benchmarking}.
}.
The terms with $k=\ell$ and $p=q$ are $\frac{u}{2N} \sum \c_{k\mu}^\dag \c_{k\mu} \c_{p\nu}^\dag \c_{p\nu} = \frac12 N u$, which is a constant  energy shift that we can freely neglect.
Defining the spin operators $\s_{\mu\nu q}\equiv \c_{q\mu}^\dag \c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians are
\begin{align}
  \H_{\t{kin}}
  &= -2J\sum_{q,\mu} \cos\p{q} \s_{\mu\mu q},
  \label{H_kin_start} \\
  \H_{\t{int}}
  &= -\f{u}{2N} \sum_{p,q,\mu,\nu} \s_{\mu\nu p} \s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by fermionic statistics (Pauli exclusion).
In this case we can simply treat our system as $N$ distinguishable $n$-level quantum spins at ``lattice sites'' $p,q$.
Note that the ``kinetic'' terms of this spin model ($\H_{\t{kin}}$) are proportional to the identity operator, contributing an overall shift in energy that we can neglect at this point.
Nevertheless, these kinetic terms will become important in the presence of an external drive, which we discuss in Section \ref{sec:SOC}.
The validity of approximating the Hubbard model in Eqs.~\eqref{eq:H_kin_FH}--\eqref{eq:H_int_FH} by the spin model in Eqs.~\eqref{H_kin_start}--\eqref{eq:H_int_start} has been previously benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}, and we provide additional benchmarking for SU(4) and SU(6) in Appendix \ref{sec:benchmarking}.

To further simplify the interaction Hamiltonian $\H_{\t{int}}$ and write it in a form reminiscent of more familiar SU(2) spin models, we now construct the operator-valued spin matrix
\begin{align}
  \bm\s_q \equiv \sum_{\mu,\nu} \s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\bm{\hat{A}},\bm{\hat{B}}$, we define the inner product
\begin{align}
  \bm{\hat{A}} \cdot \bm{\hat{B}}
  \equiv \sum_{\mu,\nu} \hat{A}_{\mu\nu}^\dag \hat{B}_{\mu\nu}.
  \label{eq:dot_product}
\end{align}
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  \H_{\t{int}} = -\f{u}{2N} \sum_{p,q} \bm\s_p\cdot\bm\s_q
  = -\f{u}{2N}\bm\S\cdot\bm\S,
  \label{eq:H_int}
\end{align}
where $\bm\S\equiv\sum_q\bm\s_q$ is a collective spin matrix, analogous to the collective spin vector $\vec S=(\S_\x,\S_\y,\S_\z)$ in the case of SU(2) \cite{he2019engineering}, with $\frac12\bm\S\cdot\bm\S \simeq \vec S\cdot\vec S = \S_\x^2 + \S_\y^2 + \S_\z^2$ when $n=2$ (here $\simeq$ denotes equality up to identity terms).

\begin{figure}
\centering
\includegraphics[width=\linewidth]{spin_model.pdf}
\caption{
(a) Ultracold atoms on a lattice of $L$ sites tunnel between neighboring lattice sites at a rate $J$, and locally repel each other with interaction energy $U$.
(b,c) When the interaction energy $U$ is small compared to the single-particle bandwidth $4J$, the frozen-mode approximation enables the interaction Hamiltonian to be written as a spin model consisting of exchange terms $\bm\s_p\cdot\bm\s_q$, which swap the states of two spins pinned to modes $p,q$.
(d) Interactions open an energy gap $u=U\times N/L$ between the manifold of permutationally symmetric states of $N$ spins, and the orthogonal complement of states that break spin-permutation symmetry.
}
\label{fig:spin_model}
\end{figure}

We now discuss the spin Hamiltonian $\H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\bm\s_p\cdot\bm\s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\bm\s_p\cdot\bm\s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\bm\s_p\cdot\bm\s_q$ is analogous to the enforcement of SU(2) spin alignment by ferromagnetic interactions, which similarly assigns different energies to the anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.
By summing over all pair-wise exchange terms $\bm\s_p\cdot\bm\s_q$, the interaction Hamiltonian $\H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $u$ between the manifold of all permutationally symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
See Figure \ref{fig:spin_model} for a summary of this section thus far.

@@

Finally, we take a moment to discuss individual $n$-level spins.
The state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of an $n$-level spin can be represented by a quasi-probability distribution $Q_\psi$ on the Bloch sphere (commonly known as the Husimi-$Q$ function, e.g.~in the spin-squeezing community \cite{ma2011quantum}).
The value $Q_\psi\p{\bm{v}}$ at a point $\bm{v}$ on the sphere is equal to the overlap of $\ket\psi$ with a pure state $\ket{\bm{v}}$ that is maximally polarized in the direction of $\bm{v}$: $Q_\psi\p{\bm{v}}\equiv\abs{\bk{\bm{v}|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\hat\rho$, this distribution is defined by $Q_{\hat\rho}\p{\bm{v}}\equiv\bk{\bm{v}|\hat\rho|\bm{v}}$.
Closely related spherical representations of multilevel spin states and operators are discussed in Refs.~\cite{dowling1994wigner, li2013weylwignermoyal}.
In practice, it is conceptually useful to identify the Hilbert space of a single $n$-level spin with the Dicke manifold of $n-1$ spin-$\frac12$ particles.

@@

We consider a specific three-laser driving scheme with a geometry sketched in Figure \ref{fig:3LD}.
Here the lattice lies in the $y$-$z$ plane at an angle $\theta$ to the $z$ axis, oriented along $\bm\ell=(0,\sin\theta,\cos\theta)$.
We set the spin quantization axis along $z$.
The laser setup consists of
\begin{enumerate*}
\item two counter-propagating right-circularly polarized lasers with drive amplitudes $\Omega_\pm$ and wavevectors $\kappa\bm{v}_\pm$, propagating in opposite directions along the $z$ axis, $\bm{v}_\pm=(0,0,\pm1)$, and
\item a third laser linearly polarized along $z$, with drive amplitude $\Omega_0$ and wavevector $\kappa \bm{v}_0$, propagating along the $x$ axis $\bm{v}_0=(1,0,0)$.
\end{enumerate*}
All driving lasers are detuned by $\Delta$ below an electronic transition.
The full Hamiltonian for this three-laser drive can be written as
\begin{align}
  \H_{\t{3LD}}^{\t{full}}
  = \sum_{j,m} \Omega_m
  \p{e^{-\i m\phi j} \s_{mj} \otimes\op{\e}{\g}_j + \t{h.c.}}
  + \Delta \hat{N}_\e,
\end{align}
where $m\in\set{+1,0,-1}$ indexes the laser pointing along $\bm{v}_m$; the SOC angle $\phi\equiv\kappa\bm{v}_+\cdot\bm\ell=\kappa \cos\theta$ (in units with lattice spacing $a=1$); $\s_{\z,j},\s_{+,j}\s_{-,j}$ are standard axial, spin-raising, and spin-lowering operators for the spin at lattice site $j$; $\s_{0,j}\equiv\s_{\z,j}$ for shorthand; $\ket\g_j$ and $\ket\e_j$ respectively denote the ground and excited electronic states of atom $j$; and $\hat{N}_\e=\1\otimes\sum_j\op\e_j$ counts the number of excited atoms (with $\1$ the identity operator on all spin degrees of freedom).

@@

We thank Victor Gurarie, Emil Yuzbashyan, Asier P.~Orioli, and Jeremy T.~Young for helpful discussions on this work.
This work was supported by the AFOSR grants FA9550-18-1-0319, FA9550-19-1-0275, by the DARPA and ARO grant W911NF-16-1-0576, ARO W911NF-19-1-0210, NSF PHY1820885, NSF JILA-PFC PHY-1734006, QLCI Award OMA - 2016244, the U.S. Department of Energy, Office of Science, National Quantum Information Science Research Centers, Quantum Systems Accelerator (QSA) and by the National Institute of Standards and Technology (NIST).
